\chapter{Przystawki}

\section{Vegan parmesan}

\begin{ingreds}
\item{almod flour, 110g, 1 kubek}
\item{oregano suszone, 2 tsp}
\item{salt, 0.5 tsp}
\item{lemon juice, 1 tbsp}
\end{ingreds}

\recipara{Preheat the oven to 180C, line with a parchment paper. Add almond flour, dried oregano and salt to a small bowl and whisk very well. Add lemon juice and use a fork to press and mix for a couple of minutes. Spread out thinly and evenly on the parchment paper. Bake just for 5 minutes until the egdes are turning gloden brown. Can mix there once or twice between putting for another 3 minutes but careful, it burns quickly.}

\commentskip

\recipend

\section{Sauce Béchamel}

\begin{ingreds}
\item{butter, 30g}
\item{flour, 30g}
\item{milk, 375g}
\item{salt, 0.25 tsp}
\end{ingreds}

\recipara{Melt butter over low heat, blend in flour and cook slowly stirring, until the butter and flour froth together for 2 min without colouring. Boil the milk with salt in another saucepan.}
\recipara{Take the roux from fire and as soon as it starts bubbling pour over all the boiling milk. Immediately beat vigorously with a whire whisk. Put on moderately high heat and stir until it comes to boil. Boil for 1 minute stirring. Add salt and pepper to taste.}

\commentskip

\comment{This is for medium thickness. For thick use 250g milk, for thin 15g flour and butter for 250g milk.}
\comment{If you have time, can give the sauce a long simmering, about 2h. Seldom necessary.}
\comment{Can enrich with cream and butter and drops of lemon juice for Sauce Suprême or that and two egg yolks for Sauce Parisienne. For than first add a little simmering sauce to the mixture of cream and yolks whisking vigorously and then the rest. Boil for 1 min on medium high whisking, then put through a fine sieve to remove coagulated bits of egg white. Lemon juice at the end.}
\comment{Add Swiss Cheese and nutmeg for Sauce Mornay, fresh tomato purée (optional parsley, chervil, basil, tarragon) for Sauce Aurore, 4 tbl of fresh or 2 tbl of dried herbs, 2 tbl shallots or spring onions boiled down in a glass of white wine or less vermuth to 2 tbl, strained away from herbs and pressed into the béchamel for Sauce Chivry (finish with enrichment butter and fresh herbs) or Sauce au Cari with 60g onions and butter cooked for 10 mins without coloring, 2 -- 3 tbl curry powder for 2 min, 45 g flour for 3 min and then milk and cream with drops of lemon juice. Sauce Soubise is the same without curry with half a kilo of onions and 90g of butter, some salt, cooked covered for 20 -- 30 mins until onions are very tender but not browned.}

\recipend

\section{Hummus}

\begin{ingreds}
\item{cieciorka, 2 kubki}
\item{soda oczyszczona, 1 łyżka}
\item{czosnek, 3--5 ząbków}
\item{cytryna, 1.5}
\item{tahina, 2.5 łyżki}
\opitem{oliwa z oliwek}
\opitem{kmin rzymski}
\opitem{wędzona papryka}
\opitem{za'atar}
\end{ingreds}

\recipara{Cieciorkę namoczyć przez 12--24 godziny.}
\recipara{Cieciorkę z czosnkiem i sodą wrzucić do garnka, zagotować i do miękkości. Nie rozgotuj, musisz obrać. Wstaw kubek wody do zamrażalnika.}
\recipara{Czosnek i 3 kubki cieciorki wrzucić do robota, sól zmieszać z sokiem z cytryny i powoli wlewać. Ciąć aż cieciorka będzie dość gładka. Dać tahiny, włączyć robota i stopniowo dolewać wody, procesować przez 4--5 min. Woda/tahina do dobrej konsystencji.}
\recipara{Wykańczać resztą cieciorki, oliwą i przyprawami.}

\commentskip

\recipend

\section{Concombres au Beurre}

\begin{ingreds}
\item{ogórki, 6}
\item{cebula szczypiorowa, 3 -- 4 łyżeczki}
\item{ocet winny, 2 łyżki}
\item{sól, 1.5 łyżeczki}
\item{cukier, 0.125 łyżeczki}
\item{masło, 50 g}
\item{bazylia lub koperek, 0.5 łyżeczki}
\end{ingreds}

\recipara{Ogórki obierz, przetnij wzdłuż, wybierz nasiona łyżką. Potnij na paluszki. Wrzuć do miski z octem, solą i cukrem na co najmniej pół godziny. Odlej, płyn, wysusz ściereczką.}
\recipara{Nagrzej piekarnik do 180 stopni, rozpuść masło. Wrzuć do żaroodpornego naczynia masło, ogórki, zioła, posiekane cebule, pieprz. Trzymaj nieprzykryte przez godzinę, przewracająć dwa, trzy razy aż będą miękkie z odribną crispiness and texture.}

\commentskip

\comment{Podawać można z pietruszką albo zagotowaną i zredukowaną do połowy śmietną z pieprzem i pietruszką albo z beszamelem z serem (sauce mornay).}
\comment{Albo smietana (250 ml), pieczarki, pietruszka, mąka ziemniaczana z wodą (łyżka each). Pieczarki obrać, umyć, poćwiartować, wysuszyć. Smażyć na średniomałym przez 5 min, wlać śmietnę i mąkę z wodą, gotować wolno przez 5 aż zgęstnieje.}

\recipend

\section{Oignons à la Grecque}

\begin{ingreds}
\item{small oignons, 0.5 kg}
\item{olive oil, 60 ml}
\item{lemon juice, 60 ml}
\item{salt, 0.5 tsp}
\item{shallots or spring onions, 2 tbl}
\item{parsley, 6 sprigs}
\item{celery stalk, 1}
\item{fennel, 1 sprig fresh or 0.125 tsp dried seeds}
\item{thyme, 1 sprig fresh or 0.125 dried}
\item{peppercorns, 12}
\item{coriander seeds, 6}
\end{ingreds}

\recipara{Prepare court bouillon, place all the ingredients in a saucepan with 0.375 ml water and simmer for 10 min.}
\recipara{Drop oignons in boiling water for 1 minute, drain and peel. Cut a cross in their root ends and cook in court bouillon for half an hour.}
\recipara{Take out the oignons, boild down the bouillon to concentrate flavour and pour over the veggies.}

\commentskip

\recipend

\section{Celéri--Rave Rémoulade}

\begin{ingreds}
\item{korzeń selera, 0.5kg}
\item{sól, 1.5 łyżeczki}
\item{sok z cytryny, 1.5 łyżeczki}
\item{musztarda francuska, 4 łyżki}
\item{oliwa, 60 -- 120 ml}
\item{ocet winny, 2 łyżki}
\opitem{świeże zioła}
\end{ingreds}

\recipara{Obierz i potnij w patyczki na tyle grube żeby zachowały sztywność. Wrzuć do miski z solą i sokiem z cytryny na 30 minut. Opłucz, wysusz ściereczką.}
\recipara{Ogrzej miskę ciepłą wodą, wrzuć musztardę i cały czas mieszając whiskiem dolej kropelka po kropelce trzy łyżki wrzącej wody, potem oliwa, ma powstać creamy, thick sauce. Tak samo z octem. Dopraw do smaku.}
\recipara{Zamarynuj na co najmniej 2,3 godziny lub na całą noc. Posyp ziołami przed podaniem.}

\commentskip

\comment{Wydaje się, że powinna to być gęsta musztarda, nie taka z widocznymi ziarnami gorczycy, z taką nie powstaje gęsty sos.}

\recipend

\section{Fasolka szparagowa po koreańsku}

\begin{ingreds}
\item{żółta fasolka szparagowa, 0.5kg}
\item{brązowy cukier, 3 łyżki}
\item{sos sojowy, 3 łyżki}
\item{mąka ziemniaczana, 1 łyżka}
\item{koncentrat pomidorowy, 1 łyżka}
\item{słodka papryka, 0.5 łyżeczki}
\item{czosnek, 2 ząbki}
\item{imbir, 1cm}
\item{pietruszka albo szczypiorek}
\opitem{sezam}
\end{ingreds}

\recipara{Fasolkę umyć, obciąc końcówki i ugotować. Odcedzić i zostawić. Czosnek przecisnąć przez praskę, imbir drobniutko posiekać.}
\recipara{Cukier rozpuścić w 0.5 szklanki ciepłej wody, wystudzić, dodać sos sojowy, przecier, paprykę i mąkę ziemniaczaną.}
\recipara{Na rozgrzanym oleju usmażyć czosnek i imbir, gdy zaczną mocno pachnieć wrzucić fasolkę i smażyć aż się zarumieni. Wlać sos, wymieszać i podgrzewać przez kilka minut aż zgętnieje. Podawać z zieleniną i prażonymi na suchej patelni ziarnami sezamu.}

\commentskip

\recipend

\section{Pureé de Marrons}

\begin{ingreds}
\item{chestnuts peeled, 1.5 l}
\item{celery, 2 stalks}
\item{parsley, 4 springs}
\item{bay leaf, 0.5}
\item{thyme, 0.125 tsp}
\item{bouillon, 0.6 l}
\end{ingreds}

\recipara{In the saucepan place the chestnuts, celery stalks and herb bouquet. pour in enough bouillon to cover the chestnuts by 1.5 inches. Simmer very slowly uncovered for 45 -- 60 min, until cooked through. Do not overcook or let them become mushy. Drain immediately, remove the celery. Purée the chestnuts.}

\commentskip

\comment{The original recipe calls for beating in butter, but I think this is undesirable.}
\comment{To peel the chestnuts: with a small sharp knife peel a strip of shell off a chestnut. Place them all in a saucepan of cold water, bring to boil and boil for 1 minute. Remove from heat. Three at a time, ladle the chestnuts out of the water and peel of the shells and inner skins. Any recalcitrant chestnuts leave for later, drop all into boiling water for a moment and peel one by one. The whole process must be done while the chestnuts are still warm.}

\recipend

